---
title: "Analyzing Diabetes SNP data"
author: "Jalal K. Siddiqui, PhD"
date: "9/27/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Introduction

This R markdown file will contain and outline code that we use to analyze CFTR SNPs on a cohort of diabetic patients and controls.  This is the final steps of the workflow.   

##### Inputting Files

The first step is inputting the data from the csv file.  This is done below.  

```{r}
diabetes.data <- read.csv("diabetes_and_cftr.csv")
head(diabetes.data)
```

The next step is to develop a binary matrix containing data whether a certain SNP is present in an SRA sample or not.  

This is done as follows

```{r}

##number of rows in diabetic data/same as total SRA samples
rows.diabetes.data <- dim(diabetes.data)[1]

##all.snps captures all snps in homozygous and heterozygous set
all.snps <- c()
for (i in 1:rows.diabetes.data){
    entry.set.homozy <- unlist(strsplit(as.character(diabetes.data$Homozygous.SNPs[i]), split = ";"))
    entry.set.heterz <- unlist(strsplit(as.character(diabetes.data$Heterozygous.SNPs[i]), split = ";"))
    all.snps <- c(all.snps, entry.set.homozy, entry.set.heterz)
}

##takes the unique SNPs
unique.snps <- unique(all.snps)

##develops and writes a matrix.snps that is binary 1 and 0 matrix describing whether a SNP is present or absent in a particular SRA sample
matrix.snps <- matrix(0, nrow = rows.diabetes.data, length(unique.snps))
rownames(matrix.snps) <- diabetes.data$SRA
colnames(matrix.snps) <- unique.snps

for (i in 1:length(unique.snps)){
   index.homoz <- grep(pattern = colnames(matrix.snps)[i], as.character(diabetes.data$Homozygous.SNPs))
   index.heter <- grep(pattern = colnames(matrix.snps)[i], as.character(diabetes.data$Heterozygous.SNPs))
   matrix.snps[index.homoz,i] <- 1
   matrix.snps[index.heter,i] <- 1
}


```

A meta object is also made that contains sample information on disease state

```{r}
meta.obj <- data.frame('name' = diabetes.data$SRA, 'Disease' = diabetes.data$Disease.State)
rownames(meta.obj) <- diabetes.data$SRA
```

##### SNP Frequency

We can see the SNP frequency as below.  

```{r}
Snp.Frequency <- colSums(matrix.snps)
hist(Snp.Frequency, breaks = 100)
```

##### Principal Component Analysis

We can do a principal component analysis of the SNP data.  This is done as follows.  

```{r}
pca.diabetes <- prcomp(matrix.snps)
library(plotly)
pca1.diab <- pca.diabetes$x[,"PC1"]
pca2.diab <- pca.diabetes$x[,"PC2"]

pca.df <- data.frame('PC1' = pca1.diab, 'PC2' = pca2.diab)
pca.df <- data.frame('name' = meta.obj$name, pca.df, 'Disease' = meta.obj$Disease)
plot_ly(data = pca.df, x = ~PC1, y = ~PC2, text = ~name, color = ~Disease, colors = c("black", "red"))
```

##### Chi-Square Test

Let's do a series of chi-square tests.

Make a function for a test matrix.  

```{r}
test.matrix.snp <- function(index.snp){
    snp.vector <- matrix.snps[,index.snp]
    snp.vector.disease <- snp.vector[diabetes.data$Disease.State == 'type ii diabetes mellitus']
    snp.vector.normal <- snp.vector[diabetes.data$Disease.State == 'normal']
    matrix.store <- matrix(0,nrow = 2, ncol = 2)
    rownames(matrix.store) <- c("snp+", "snp-")
    colnames(matrix.store) <- c("disease", "normal")
    matrix.store[1,1] <- sum(snp.vector.disease)
    matrix.store[2,1] <- length(snp.vector.disease) - sum(snp.vector.disease)
    matrix.store[1,2] <- sum(snp.vector.normal)
    matrix.store[2,2] <- length(snp.vector.normal) - sum(snp.vector.normal)
    test.matrix.snp <- matrix.store
    return(test.matrix.snp)
    
}
```

Do some sanity checks on the test matrices

```{r}
sum.test <- c()
for (i in 1:length(unique.snps)){
    sum.test[i] <- sum(test.matrix.snp(i))
}

disease.count <- c()
for (i in 1:length(unique.snps)){
    disease.count[i] <- colSums(test.matrix.snp(i))[1]
}

normal.count <- c()
for (i in 1:length(unique.snps)){
    normal.count[i] <- colSums(test.matrix.snp(i))[2]
}

```


The chi-square test is run as follows.  

```{r}
p.value.list <- c()

for (i in 1:length(unique.snps)){
obj <- chisq.test(test.matrix.snp(i)+1)
p.value.list <- c(p.value.list, obj$p.value)
}

names(p.value.list) <- as.character(unique.snps)

hist(p.value.list)

p.adjust.list <- p.adjust(p.value.list, method = "bonferroni")

hist(p.adjust.list)
```

##### Fisher's Exact Test

We can run Fisher's Exact test.  

```{r}
p.value.list.fisher <- c()

for (i in 1:length(unique.snps)){
obj <- fisher.test(test.matrix.snp(i))
p.value.list.fisher <- c(p.value.list.fisher, obj$p.value)
}

names(p.value.list.fisher) <- as.character(unique.snps)

hist(p.value.list.fisher)

p.adjust.list.fisher <- p.adjust(p.value.list.fisher, method = "bonferroni")

hist(p.adjust.list.fisher)
```

##### Incorporating More Sample Data

