---
title: "Analyzing Diabetes SNP data-Condensed"
author: "Jalal K. Siddiqui, PhD"
date: "9/27/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Introduction

This R markdown file will contain and outline code that we use to analyze CFTR SNPs on a cohort of diabetic patients and controls.  This is the final steps of the workflow.   

##### Inputting Files

The first step is inputting the data from the csv file.  This is done below.  We have 3514 samples from SRA.

```{r}
rm(list = ls())
diabetes.data <- read.csv("new_diabetes_and_cftr.csv")
head(diabetes.data)
dim(diabetes.data)
```

The next step is to develop a binary matrix containing data whether a certain SNP is present in an SRA sample or not.  

This is done as follows

```{r}

##number of rows in diabetic data/same as total SRA samples
rows.diabetes.data <- dim(diabetes.data)[1]

##all.snps captures all snps in homozygous and heterozygous set
all.snps <- c()
for (i in 1:rows.diabetes.data){
    entry.set.homozy <- unlist(strsplit(as.character(diabetes.data$Homozygous.SNPs[i]), split = ";"))
    entry.set.heterz <- unlist(strsplit(as.character(diabetes.data$Heterozygous.SNPs[i]), split = ";"))
    all.snps <- c(all.snps, entry.set.homozy, entry.set.heterz)
}

##takes the unique SNPs
unique.snps <- unique(all.snps)

##develops and writes a matrix.snps that is binary 1 and 0 matrix describing whether a SNP is present or absent in a particular SRA sample
matrix.snps <- matrix(0, nrow = rows.diabetes.data, length(unique.snps))
rownames(matrix.snps) <- diabetes.data$SRA
colnames(matrix.snps) <- unique.snps

for (i in 1:length(unique.snps)){
   index.homoz <- grep(pattern = colnames(matrix.snps)[i], as.character(diabetes.data$Homozygous.SNPs))
   index.heter <- grep(pattern = colnames(matrix.snps)[i], as.character(diabetes.data$Heterozygous.SNPs))
   matrix.snps[index.homoz,i] <- 1
   matrix.snps[index.heter,i] <- 1
}

dim(matrix.snps)
```

We have 2514 samples with 189 SNPs

A meta object is also made that contains sample information on disease state

```{r}
meta.obj <- data.frame('name' = diabetes.data$SRA, 'Disease' = diabetes.data$Disease.State)
rownames(meta.obj) <- diabetes.data$SRA
```


##### Incorporating More Sample Data

Given that these individual SRA ids are not patient data but sample data we need more information detailing information on each of these samples.  

```{r}
add.samp.meta.read <- read.table(file = "E-MTAB-5061.sdrf.txt", sep = "\t")
add.samp.meta <- add.samp.meta.read[2:dim(add.samp.meta.read)[1],]

for (i in 1:39){
colnames(add.samp.meta)[i] <- as.character(add.samp.meta.read[1,i])
}
rownames(add.samp.meta) <- add.samp.meta$`Comment[ENA_RUN]`

add.samp.meta <- data.frame(add.samp.meta, 'name' = rownames(add.samp.meta))

samp.total.meta <- add.samp.meta[rownames(meta.obj),]
dim(samp.total.meta)
```

We want to redo this analysis by only taking cell samples that are of decent quality

```{r}
good.quality.list <- which(samp.total.meta$Characteristics.single.cell.well.quality == 'OK')
length(good.quality.list)

```

From 3512 samples, only 2209 are of good quality.  

Let's get a matrix and meta object with only good quality samples.  

```{r}
meta.quality <- samp.total.meta[good.quality.list,]
matrix.quality.snps <- matrix.snps[good.quality.list,]
```

##### PCA by Cell Types

```{r}
pca.quality.diabetes <- prcomp(matrix.quality.snps)
library(plotly)
pca1.diab.q <- pca.quality.diabetes$x[,"PC1"]
pca2.diab.q <- pca.quality.diabetes$x[,"PC2"]

pca.df.q <- data.frame('PC1' = pca1.diab.q, 'PC2' = pca2.diab.q)
pca.df.q <- data.frame('name' = meta.quality$name, pca.df.q, 'Cell' = meta.quality$Characteristics.cell.type.)
plot_ly(data = pca.df.q, x = ~PC1, y = ~PC2, text = ~name, color = ~Cell, colors = c("black", "blue", "red", "green", "yellow", "orange", "magenta", "gray", "brown", "purple", "pink", "gold", "darkred", "darkgreen", "darkblue", "darkorange"))
```


##### Ductal Cells

We will do the PCA on the ductal cells alone.  First make the matrix and meta object for this information

```{r}
ductal.list <- which(meta.quality$Characteristics.cell.type. == "ductal cell")
meta.quality.d <- meta.quality[ductal.list,]
matrix.quality.snps.d <- matrix.quality.snps[ductal.list,]
```

Then do the PCA

```{r}
pca.quality.diabetes.d <- prcomp(matrix.quality.snps.d)
library(plotly)
pca1.diab.q.d <- pca.quality.diabetes.d$x[,"PC1"]
pca2.diab.q.d <- pca.quality.diabetes.d$x[,"PC2"]

pca.df.q.d <- data.frame('PC1' = pca1.diab.q.d, 'PC2' = pca2.diab.q.d)
pca.df.q.d <- data.frame('name' = meta.quality.d$name, pca.df.q.d, 'Cell' = meta.quality.d$Characteristics.cell.type., 'Patient' = meta.quality.d$Characteristics.individual., 'Disease' = meta.quality.d$Characteristics.disease.)
plot_ly(data = pca.df.q.d, x = ~PC1, y = ~PC2, text = ~name, color = ~Patient, colors = c("black", "blue", "red", "green", "yellow", "orange", "magenta", "gray", "brown", "purple", "pink", "gold", "darkred", "darkgreen", "darkblue", "darkorange"), symbol = ~Disease, symbols = c("circle", "o"))
```

You are able to analyze how different people are able to express gene-specific variants in real time.  You can see whether an individual's ductal cells are expressing variants of CFTR gene.  

